---
title: "R for the Real World: Counting those Business Days like a Pro!"
author: "Steven P. Sanderson II, MPH"
date: "2024-02-01"
categories: [code, rtip, timeseries]
---

# Introduction

Hi fellow coders, data wranglers, and all-around R enthusiasts! Have you ever been stuck calculating the number of business days between two dates? You know, like figuring out how long that project _actually_ took, excluding weekends (because let's be honest, who works on those?). Well, fret no more! Today, we're diving into the wonderful world of _business day calculations_ in R with some easy-to-follow examples. Buckle up, it's gonna be a productive ride!

# Examples 

## Step 1: Grabbing the Toolkit

First things first, we need the right tools. We'll be using the mighty _bizdays_ package. Think of it as your personal business day calculator, always ready to lend a hand (or rather, some code). Install it with this magic spell:

```{r message=FALSE, warning=FALSE}
# install.packages("bizdays")
library(bizdays)
```

## Step 2: The Basic Count

Alright, let's say you want to know how many business days there were between January 1st and December 31st, 2023. Simple, right? Here's the code:

```{r}
start_date <- as.Date("2023-01-01")
end_date <- as.Date("2023-12-31")

business_days <- bizdays(start_date, end_date, "weekends")

print(paste0("There were ", business_days, " business days in 2023!"))
```

What's happening here? We define the start and end dates, feed them to the `bizdays` function, and voila! It counts the business days for us, excluding weekends by default. The `print` function just displays the result with a fun message.

## Step 3: Get Creative and Explore!

Remember, this is just the tip of the `bizdays` iceberg. You can explore its other features like:

* Adding or subtracting business days from a date
* Handling custom holiday lists
* Working with different time zones

_But wait, there's more!_ The most important step is to **experiment and try things out yourself**. Play with different dates, holidays, and weekend definitions. See what results you get and how they fit your specific needs. R is all about exploration and making it work for you!

So, fellow coders, go forth and conquer those business day calculations with confidence! And if you get stuck, remember, the R community is always here to help. Happy coding!