---
title: "Data Preppers with {healthyR.ai}"
author: "Steven P. Sanderson II, MPH"
date: "2023-02-24"
categories: [code, rtip, healthyrai, preprocessor]
---

# Introduction

There are many different methods that one can choose from in order to model their data. This brings with it a fundamental issue of how to prepare your data for the specified algorithm. With the [`{healthyR.ai}`] package there are many different functions in this family that will help solve this issue for some algorithms but of course not all, that would be utterly exhausting for me to do on my own.

In __healthyR.ai__ I call these _Data Preppers_ because they prep the data you supply to the format necessary for the algorithm to function properly.

Let's take a look at one.

# Function

Here we are going to use the `hai_c50_data_prepper(.data, .recipe_formula)` function.

```{r eval=FALSE}
hai_c50_data_prepper(.data, .recipe_formula)
```

Here are the simple arguments:

*  `.data` - The data that you are passing to the function. Can be any type of data that is accepted by the data parameter of the recipes::recipe() function.
*  `.recipe_formula` - The formula that is going to be passed. For example if you are using the iris data then the formula would most likely be something like `Species ~` .


# Example

Here is a small example:

```{r warning=FALSE, message=FALSE}
library(healthyR.ai)

hai_c50_data_prepper(.data = Titanic, .recipe_formula = Survived ~ .)

rec_obj <- hai_c50_data_prepper(Titanic, Survived ~ .)
get_juiced_data(rec_obj)
```

Here are the rest of the data-preppers at the time of writing this article:

*  [hai_c50_data_prepper()](https://www.spsanderson.com/healthyR.ai/reference/hai_c50_data_prepper.html)
*  [hai_cubist_data_prepper()](https://www.spsanderson.com/healthyR.ai/reference/hai_cubist_data_prepper.html)
*  [hai_earth_data_prepper()](https://www.spsanderson.com/healthyR.ai/reference/hai_earth_data_prepper.html)
*  [hai_glmnet_data_prepper()](https://www.spsanderson.com/healthyR.ai/reference/hai_glmnet_data_prepper.html)
*  [hai_knn_data_prepper()](https://www.spsanderson.com/healthyR.ai/reference/hai_knn_data_prepper.html)
*  [hai_ranger_data_prepper()](https://www.spsanderson.com/healthyR.ai/reference/hai_ranger_data_prepper.html)
*  [hai_svm_poly_data_prepper()](https://www.spsanderson.com/healthyR.ai/reference/hai_svm_poly_data_prepper.html)
*  [hai_svm_rbf_data_prepper()](https://www.spsanderson.com/healthyR.ai/reference/hai_svm_rbf_data_prepper.html)
*  [hai_xgboost_data_prepper()](https://www.spsanderson.com/healthyR.ai/reference/hai_xgboost_data_prepper.html)

Voila!