---
title: "Quickly Generate Nested Time Series Models"
author: "Steven P. Sanderson II, MPH"
date: "2023-02-27"
categories: [code, rtip, timeseries, autoarima]
---

# Introduction

There are many approaches to modeling time series data in R. One of the types of data that we might come across is a nested time series. This means the data is grouped simply by one or more keys. There are many methods in which to accomplish this task. This will be a quick post, but if you want a longer more detailed and quite frankly well written out one, then [this is a really good article](https://rpubs.com/alfado/purrr)

# Exampmle

Let's just get to it with a very simple example, the motivation here isn't to be all encompassing, but rather to just showcase it is possible for those who may not know it is.

```{r, warning=FALSE, message=FALSE}
library(healthyR.data)
library(dplyr)
library(timetk)

ts_tbl <- healthyR_data |> 
  filter(ip_op_flag == "I") |> 
  select(visit_end_date_time, service_line, length_of_stay) |>
  mutate(visit_end_date_time = as.Date(visit_end_date_time)) |>
  group_by(service_line) |>
  summarise_by_time(
    .date_var = visit_end_date_time,
    .by = "month",
    los = mean(length_of_stay)
  ) |>
  ungroup()

glimpse(ts_tbl)
```

```{r warning=FALSE, message=FALSE}
library(forecast)
library(broom)
library(tidyr)

glanced_models <- ts_tbl |> 
  nest_by(service_line) |> 
  mutate(AA = list(auto.arima(data$los))) |> 
  mutate(perf = list(glance(AA))) |> 
  unnest(cols = c(perf))

glanced_models |>
  select(-data)
```

Voila!
