---
title: "New Release of {healthyR.ts}"
author: "Steven P. Sanderson II, MPH"
date: "2023-01-09"
categories: [code, rtip, healthyrts]
---

# Introduction

Hello R users!

I am are excited to announce a new update to the [`{healthyR.ts}`](https://www.spsanderson.com/healthyR.ts/) package: the [`ts_brownian_motion()`](https://www.spsanderson.com/healthyR.ts/reference/ts_brownian_motion.html) function.

This function allows you to easily simulate brownian motion, also known as a Wiener process, using just a few parameters. You can specify the length of the simulation using the '.time' parameter, the number of simulations to run using the '.num_sims' parameter, the time step size (standard deviation) using the '.delta_time' parameter, and the initial value (which is set to 0 by default) using the '.initial_value' parameter.

But what is brownian motion, and why might you want to simulate it? Brownian motion is a random process that describes the movement of particles suspended in a fluid. It is named after the botanist Robert Brown, who observed the random movement of pollen grains suspended in water under a microscope in the 19th century.

In finance, brownian motion is often used to model the movement of stock prices over time. By simulating brownian motion, you can get a sense of how prices might fluctuate in the future, and use this information to inform your investment decisions.

I hope that the `ts_brownian_motion()` function will be a useful tool for anyone interested in simulating brownian motion, whether for financial modeling or any other application. Give it a try and see what you can do with it!

Right now the function is a bit slow at `.num_sims` > 500 so I am working on optimizing it. I will also later on be introducing the Geometric Brownian Motion to `{healthyR.ts}`

As always, we welcome feedback and suggestions for new features and improvements. Thank you for using the {healthyR.ts} package, and happy simulating!

# Function

Here is the full function call:

```{r eval=FALSE}
ts_brownian_motion(
  .time = 100,
  .num_sims = 10,
  .delta_time = 1,
  .initial_value = 0
)
```

# Example

A simple example of the output.

```{r warning=FALSE, message=FALSE}
library(healthyR.ts)

ts_brownian_motion()
```

Voila!