---
title: "Mastering the Art of Drawing Circles in Plots with R"
author: "Steven P. Sanderson II, MPH"
date: "2023-10-13"
categories: [rtip, viz]
---

# Introduction

As an R programmer, you may want to draw circles in plots to highlight certain data points or to create visualizations. Here are some simple steps to draw circles in plots using R:

# Examples

1. First, create a scatter plot using the `plot()` function in R. For example, you can create a scatter plot of `x` and `y` values using the following code:

```{r}
x <- c(1, 2, 3, 4, 5)
y <- c(2, 4, 6, 8, 10)
plot(x, y)
```

2. To draw a circle on the plot, you can use the `symbols()` function in R. The `symbols()` function allows you to draw various shapes, including circles, squares, triangles, and more. To draw a circle, set the `circles` argument to `TRUE`. For example, to draw a circle with a radius of 0.5 at the point (3, 6), use the following code:

```{r}
plot(x, y)
symbols(3, 6, circles = 1, add = TRUE)
```

3. You can also customize the color and border of the circle using the `bg` and `fg` arguments. For example, to draw a red circle with a blue border, use the following code:

```{r}
plot(x, y)
symbols(3, 6, circles = 1, add = TRUE, bg = "red", fg = "blue")
```

4. To draw multiple circles on the plot, you can use a loop to iterate over a list of coordinates and radii. For example, to draw three circles with different radii at different points, use the following code:

```{r}
plot(x, y)

coords <- list(c(2, 4), c(3, 6), c(4, 8))
radii <- c(0.1, 0.2, 0.3)

for (i in 1:length(coords)) {
  symbols(
    coords[[i]][1], coords[[i]][2], circles = radii[[i]], 
    add = TRUE, bg = "red", fg = "blue", inches = FALSE
  )
}
```

5. Finally, you can add a title and axis labels to the plot using the `title()`, `xlab()`, and `ylab()` functions. For example, to add a title and axis labels to the plot, use the following code:

```{r}
x <- c(1, 2, 3, 4, 5)
y <- c(2, 4, 6, 8, 10)
plot(
  x, y, main = "Scatter Plot with Circles", 
  xlab = "X Values", ylab = "Y Values"
)

coords <- list(c(2, 4), c(3, 6), c(4, 8))
radii <- c(0.1, 0.2, 0.3)

for (i in 1:length(coords)) {
  symbols(
    coords[[i]][1], coords[[i]][2], circles = radii[[i]], 
    add = TRUE, bg = "red", fg = "blue", inches = FALSE
  )
}
```
Here is one last exmple:

```{r}
# Create a scatter plot with multiple circles
n <- 10
x <- runif(n, -2, 2)
y <- runif(n, -2, 2)
size <- runif(n, 0.1, 1)
fill <- sample(colors(), n)
border <- sample(colors(), n)

symbols(x, y, circles = size, inches = FALSE, add = F, bg = fill, fg = border)

```
# Conclusion

Overall, drawing circles in plots is a simple and effective way to highlight certain data points or to create visualizations. Try experimenting with different coordinates, radii, colors, and borders to create your own custom plots.