---
title: "Why Check File Size Output for Different Methods?"
author: "Steven P. Sanderson II, MPH"
date: "2023-05-26"
categories: [rtip, excel, openxlsx, xlsx, writexl]
---

# Introduction

When working with data, it is important to be aware of the file size of the data you are working with. This is especially true when you are working with large datasets, as the file size can have a significant impact on the performance of your code.

In R, there are a number of different ways to write data to files. Each method has its own advantages and disadvantages, and the file size of the output can vary depending on the method you use.

In this blog post, we will discuss why it is a good idea to check the file size output for different methods. We will also provide three examples of how to check the file size output using the R libraries writexl, openxlsx, and xlsx.

# Why Check File Size Output?

There are a number of reasons why it is a good idea to check the file size output for different methods.

*    To ensure that the data is being written correctly: If the file size of the output is significantly different from the expected file size, it is a sign that something may have gone wrong when the data was being written to the file. This could be due to a number of factors, such as a typo in the code, a problem with the file system, or a bug in the R library.
*    To optimize performance: The file size of the output can have a significant impact on the performance of your code. If the file size is too large, it can take longer to read the data from the file, which can slow down your code.
*    To troubleshoot problems: If you are having problems with your code, checking the file size output can help you to identify the source of the problem. For example, if the file size of the output is significantly smaller than the expected file size, it is a sign that the data may not be being written to the file correctly.

# Examples of Checking File Size Output

In R, there are a number of different ways to check the file size output. Here are three examples of how to check the file size output using the R libraries writexl, openxlsx, and xlsx:

## `writexl`

To check the file size output of the `writexl::write_xlsx()` function, you can use the `file.info()` function. For example, the following code will write the iris dataset to a temporary file and then print the file size of the 

```{r message=FALSE, warning=FALSE}
library(writexl)

write_xlsx(iris, tmp1 <- tempfile())

file.info(tmp1)$size
```

## `openxlsx`

To check the file size output of the `openxlsx::write.xlsx()` function, you can use the `file.info()` function. For example, the following code will write the iris dataset to a temporary file and then print the file size of the output:

```{r message=FALSE, warning=FALSE}
library(openxlsx)

write.xlsx(iris, tmp2 <- tempfile())

file.info(tmp2)$size
```

## `xlsx`

To check the file size output of the `xlsx::write.xlsx()` function, you can use the file.info() function. For example, the following code will write the iris dataset to a temporary file and then print the file size of the output:

```{r message=FALSE, warning=FALSE}
library(xlsx)

write.xlsx(iris, tmp3 <- paste0(tempfile(), ".xlsx"))

file.info(tmp3)$size
```

# Conclusion

In this blog post, we discussed why it is a good idea to check the file size output for different methods. We also provided three examples of how to check the file size output using the R libraries writexl, openxlsx, and xlsx.

By checking the file size output, you can ensure that the data is being written correctly, optimize the performance of your code, and troubleshoot problems.