---
title: "Solving Systems of Equations in R using the `solve()` Function"
author: "Steven P. Sanderson II, MPH"
date: "2023-08-15"
categories: [rtip, linearequations]
---

# Introduction

In mathematical modeling and data analysis, it is often necessary to solve systems of equations to find the values of unknown variables. R provides the `solve()` function, which is a powerful tool for solving systems of linear equations. In this blog post, we will explore the purpose of solving systems of equations, explain the syntax of the `solve()` function, and provide three examples of increasing complexity to demonstrate its usage.

# Purpose of Solving Systems of Equations

Solving systems of equations allows us to find the values of unknown variables that satisfy multiple equations simultaneously. This is useful in various fields, including physics, engineering, economics, and data analysis. By solving systems of equations, we can determine the relationships between variables and make informed decisions based on the solutions obtained.

# Syntax of the `solve()` Function

The `solve()` function in R is used to solve linear algebraic equations of the form "a %*% x = b", where "a" is the coefficient matrix, "x" is the vector or matrix of unknown variables, and "b" is the vector or matrix of constants. The `solve()` function takes two arguments: "a" and "b".

# Examples

## Example 1 Solving a System of Two Equations

Let's start with a simple example of solving a system of two equations with two variables. Suppose we have the following system of equations:

```
2x + 3y = 10
4x - 2y = 6
```

To solve this system using the `solve()` function, we define the coefficient matrix "a" and the constant matrix "b" as follows:

```{r}
a <- matrix(c(2, 3, 4, -2), nrow = 2, byrow = TRUE)
b <- c(10, 6)
```

Then, we can use the `solve()` function to find the values of "x" and "y":

```{r}
solution <- solve(a, b)
solution
```

The solution will be stored in the "solution" variable, which can be accessed to obtain the values of "x" and "y".

## Example 2 Solving a System of Three Equations

Let's consider a slightly more complex system of three equations with three variables:

```R
3x + 2y - z = 7
x - y + 2z = -1
2x + 3y + 4z = 12
```

To solve this system, we define the coefficient matrix "a" and the constant matrix "b":

```{r}
a <- matrix(c(3, 2, -1, 1, -1, 2, 2, 3, 4), nrow = 3, byrow = TRUE)
b <- c(7, -1, 12)
```

We can then use the `solve()` function to find the values of "x", "y", and "z":

```{r}
solution <- solve(a, b)
solution
```

The solution will be stored in the "solution" variable, and we can access the values of "x", "y", and "z" from it.

## Example 3 Solving a System of Equations with Matrix Coefficients
In some cases, the coefficient matrix "a" can be a matrix instead of a vector. For example, consider the following system of equations:

```
2x + 3y = 10
4x - 2y = 6
```

We can represent the coefficient matrix "a" as follows:

```{r}
a <- matrix(c(2, 3, 4, -2), nrow = 2, byrow = TRUE)
```

The constant vector "b" remains the same:

```{r}
b <- c(10, 6)
```

We can then use the `solve()` function to find the values of "x" and "y":

```{r}
solution <- solve(a, b)
solution
```

The solution will be stored in the "solution" variable, and we can access the values of "x" and "y" from it.

# Conclusion

In this blog post, we explored the `solve()` function in R, which is a powerful tool for solving systems of equations. We discussed the purpose of solving systems of equations and explained the syntax of the `solve()` function. We provided three examples of increasing complexity to demonstrate how to use the `solve()` function to solve systems of equations. We encourage readers to try solving their own systems of equations using the `solve()` function in R to gain a deeper understanding of its capabilities.

# Resources

*   https://www.geeksforgeeks.org/solve-system-of-equations-in-r/
*   https://statisticsglobe.com/solve-system-of-equations-in-r/
*   https://sparkbyexamples.com/r-programming/r-solve-equation/
*   https://www.statology.org/solve-system-of-equations-in-r/
*   https://youtube.com/watch?v=EZhUNhSIomE
*   https://youtube.com/watch?v=Z0evGDOrv5w