---
title: "Unveiling the Power of `get_cms_meta_data()` in healthyR.data"
author: "Steven P. Sanderson II, MPH"
date: "2024-05-28"
categories: [code, rtip, healthyrdata]
toc: TRUE
---

# Introduction

Hey, R users! 🌟 Today, we're going to look at a great new addition to the **healthyR.data** package—the `get_cms_meta_data()` function! This function is a helpful tool for retrieving and analyzing metadata from CMS (Centers for Medicare & Medicaid Services) datasets. Whether you're a healthcare analyst, data scientist, or R programming fan, you'll find this function very useful. Let's break it down and explore how it works.

## Overview of `get_cms_meta_data()`

The `get_cms_meta_data()` function lets you retrieve metadata from CMS datasets easily. You can customize your search using various parameters, ensuring you get precisely the data you need. Here's the syntax:

```R
get_cms_meta_data(
  .title = NULL,
  .modified_date = NULL,
  .keyword = NULL,
  .identifier = NULL,
  .data_version = "current",
  .media_type = "all"
)
```

### Arguments:

-   `.title`: Search by title.
-   `.modified_date`: Search by modified date (format: "YYYY-MM-DD").
-   `.keyword`: Search by keyword.
-   `.identifier`: Search by identifier.
-   `.data_version`: Choose between "current", "archive", or "all". Default is "current".
-   `.media_type`: Filter by media type ("all", "csv", "API", "other"). Default is "all".

### Return Value:

A tibble containing data links and relevant metadata about the datasets.

### Details:

The function fetches JSON data from the CMS data URL and extracts relevant fields to create a tidy tibble. It selects specific columns, handles nested lists by unnesting them, cleans column names, and processes dates and media types to make the data more useful for analysis. The columns in the returned tibble include:

-   `title`
-   `description`
-   `landing_page`
-   `modified`
-   `keyword`
-   `described_by`
-   `fn`
-   `has_email`
-   `identifier`
-   `start`
-   `end`
-   `references`
-   `distribution_description`
-   `distribution_title`
-   `distribution_modified`
-   `distribution_start`
-   `distribution_end`
-   `media_type`
-   `data_link`

## Practical Examples

Let's see the `get_cms_meta_data()` function in action with a couple of examples.

### Example 1: Basic Usage

First, we'll load the necessary libraries and fetch some metadata:

```{r message=FALSE, warning=FALSE}
# Library Loads
library(healthyR.data)
library(dplyr)

# Get data
cms_data <- get_cms_meta_data()
glimpse(cms_data)

# Attributes
atb <- attributes(cms_data)
atb$names
atb$class
atb$url
atb$date_retrieved
atb$parameters
```

In this example, we're simply calling `get_cms_meta_data()` without any parameters. This fetches the default dataset metadata. The `glimpse()` function from the `dplyr` package provides a quick overview of the data structure.

### Example 2: Custom Search by Keyword and Title

Now, let's refine our search by specifying a keyword and title:

```{r message=FALSE, warning=FALSE}
get_cms_meta_data(
  .keyword = "nation",
  .title = "Market Saturation & Utilization State-County"
) |>
  glimpse()
```

In this example, we filter the metadata by the keyword "nation" and the title "Market Saturation & Utilization State-County". The pipe operator (`|>`) is used to pass the result directly into the `glimpse()` function for a quick preview.

## Breaking Down the Code

Let's break down the code blocks to understand what they're doing:

### Basic Usage

1. **Load Libraries**: 
   ```R
   library(healthyR.data)
   library(dplyr)
   ```
   We load the `healthyR.data` package to access the `get_cms_meta_data()` function and the `dplyr` package for data manipulation.

2. **Fetch Metadata**:
   ```R
   cms_data <- get_cms_meta_data()
   ```
   We call `get_cms_meta_data()` without any parameters to get the default dataset metadata.

3. **Preview Data**:
   ```R
   glimpse(cms_data)
   ```
   The `glimpse()` function gives us a quick look at the structure and contents of the fetched metadata.

### Custom Search

1. **Custom Search Call**:
   ```R
   get_cms_meta_data(
     .keyword = "nation",
     .title = "Market Saturation & Utilization State-County"
   ) |>
   glimpse()
   ```
   Here, we call `get_cms_meta_data()` with specific parameters for keyword and title to narrow down our search. The result is passed to `glimpse()` using the pipe operator for an immediate preview.

## Conclusion

The `get_cms_meta_data()` function is a versatile and flexible tool for accessing CMS metadata, making your data analysis tasks more efficient and effective. Whether you're looking for specific datasets or just exploring the available metadata, this function has got you covered.

Try out `get_cms_meta_data()` in your next R project and explore the potential of CMS data with ease! Happy coding! 🚀