---
title: "Checking if a String Contains Multiple Substrings in R"
author: "Steven P. Sanderson II, MPH"
date: "2024-07-23"
categories: [code, rtip, operations, strings]
toc: TRUE
---

# Introduction

Hello, fellow R programmers! Today, we're looking at a practical topic that often comes up when dealing with text data: how to check if a string contains multiple substrings. We'll cover how to do this in base R, as well as using the `stringr` and `stringi` packages. Each approach has its own advantages, so let's explore them together.

# Examples

## Base R Approach

First, let's start with base R. Suppose we have a string and we want to check if it contains both "apple" and "banana". Here's how you can do it:

```{r}
# Our main string
main_string <- "I have an apple and a banana."

# Substrings to check
substrings <- c("apple", "banana")

# Check if all substrings are in the main string
contains_all <- all(sapply(substrings, function(x) grepl(x, main_string)))

# Output the result
contains_all

sapply(substrings, grepl, x = main_string)
```

### Explanation

1.    **`main_string`**: This is the string we are checking.
2.    **`substrings`**: A vector containing the substrings we are looking for.
3.    **`sapply(substrings, function(x) grepl(x, main_string))`**: We use `sapply` to apply `grepl` (which checks if a pattern is found in a string) to each substring. This returns a logical vector indicating if each substring is present.
4.    **`all()`**: This function checks if all values in the logical vector are `TRUE`.

By combining these functions, we can efficiently check if all the substrings are present in our main string.

## Using `stringr`

The `stringr` package provides a set of functions designed to make string manipulation easier and more intuitive. Here's how we can use it to achieve the same goal:

```{r message=FALSE, warning=FALSE}
# Load the stringr package
library(stringr)

# Our main string
main_string <- "I have an apple and a banana."

# Substrings to check
substrings <- c("apple", "banana")

# Check if all substrings are in the main string
contains_all <- all(str_detect(main_string, substrings))

# Output the result
contains_all

str_detect(main_string, substrings)
```

### Explanation

1.    **`library(stringr)`**: Loads the `stringr` package.
2.    **`str_detect(main_string, substrings)`**: The `str_detect` function checks if each pattern in `substrings` is found in `main_string`. It returns a logical vector.
3.    **`all()`**: As before, `all` checks if all values in the logical vector are `TRUE`.

The `stringr` package simplifies the syntax and makes the code more readable.

## Using `stringi`

The `stringi` package is another powerful tool for string manipulation. It offers a highly efficient way to handle strings. Here's how we can use `stringi` to check for multiple substrings:

```{r message=FALSE, warning=FALSE}
# Load the stringi package
library(stringi)

# Our main string
main_string <- "I have an apple and a banana."

# Substrings to check
substrings <- c("apple", "banana")

# Check if all substrings are in the main string
contains_all <- all(stri_detect_fixed(main_string, substrings))

# Output the result
contains_all

stri_detect_fixed(main_string, substrings)
```

### Explanation

1.    **`library(stringi)`**: Loads the `stringi` package.
2.    **`stri_detect_fixed(main_string, substrings)`**: The `stri_detect_fixed` function checks if each fixed pattern in `substrings` is found in `main_string`. This function is optimized for fixed patterns and is very fast.
3.    **`all()`**: Again, we use `all` to check if all values in the logical vector are `TRUE`.

`stringi` provides highly optimized functions that can be very useful for handling large datasets or performance-critical applications.

# Try It Yourself!

Now that we've walked through the different methods to check if a string contains multiple substrings, I encourage you to try these approaches on your own. Experiment with different strings and substrings to get a feel for how these functions work. Understanding these techniques can greatly enhance your text data manipulation skills in R.

Happy coding, and feel free to share your experiences and any questions you might have in the comments!