---
title: "Reading in Multiple Excel Sheets with lapply and {readxl}"
author: "Steven P. Sanderson II, MPH"
date: "2023-04-07"
categories: [rtip, readxl, lapply, healthyr]
---

# Intruduction

Reading in an Excel file with multiple sheets can be a daunting task, especially for users who are not familiar with the process. In this blog post, we will walk through a sample function that can be used to read in an Excel file with multiple sheets using the R programming language.

# Function

The function we will be using is called `excel_sheet_reader()`. This function takes one argument: `filename`, which is the name of the Excel file we want to read in. This function, since it is using the `{readxl}` package will automatically read that data to a tibble.

# Example

Here is the function:

```{r}
excel_sheet_reader <- function(filename) {
  sheets <- excel_sheets(filename)
  x <- lapply(sheets, function(X) read_excel(filename, sheet = X))
  names(x) <- sheets
  x
}
```

The first thing the `excel_sheet_reader()` function does is to determine the names of all the sheets in the Excel file using the excel_sheets function from the readxl package. This function returns a character vector containing the names of all the sheets in the Excel file.

```{r eval=FALSE}
sheets <- excel_sheets(filename)
```

Next, the function uses the `lapply` function to loop through all the sheet names and read in each sheet using the `read_excel()` function, also from the `readxl` package. This function takes two arguments: filename, which is the name of the Excel file, and sheet, which is the name of the sheet we want to read in. The `lapply` function returns a list containing all the sheets.

```{r eval=FALSE}
x <- lapply(sheets, function(X) read_excel(filename, sheet = X))
```

Finally, the function uses the names function to assign the sheet names to the list of sheets and returns the list.

```{r eval=FALSE}
names(x) <- sheets
x
```

Now that we have explained the `excel_sheet_reader()` function, let's use it to read in the iris and mtcars datasets.

```{r warning=FALSE, message=FALSE}
library(healthyR)
library(dplyr)
library(writexl)
library(readxl)

iris |>
  named_item_list(Species) |>
  write_xlsx(path = "iris.xlsx")

mtcars |>
  named_item_list(cyl) |>
  write_xlsx(path = "mtcars.xlsx")

iris_sheets <- excel_sheet_reader("iris.xlsx")
mtcars_sheets <- excel_sheet_reader("mtcars.xlsx")
```

Now lets see the structure of each file.

```{r}
iris_sheets
```

Now `mtcars_sheets`

```{r}
mtcars_sheets
```

And that's it! Hope this has been helpful!
