---
title: "Cumulative Skewness"
author: "Steven P. Sanderson II, MPH"
date: "2022-10-31"
categories: [code, rtip]
---

# Function

In this post we will make a function `cum_skewness()` that will generate a vector output of the cumulative skewness of some given vector. The full function call is simply:

```{r fns, eval=FALSE}
cum_skewness(.x)
```

It only takes in a numeric vector, we are not going to write type checks in the function as it won't be necessary for this post.

```{r fns_c}
cum_skewness <- function(.x){
  skewness <- function(.x){
    sqrt(length(.x)) * sum((.x - mean(.x))^3 / (sum((.x))^2)^(3/2))
  }
  sapply(seq_along(.x), function(k, z) skewness(z[1:k]), z = .x)
}
```

# Data

We are going to use the `mtcars` data set and use the `mpg` column for this example. Let's set `x` equal to `mtcars$mpg`

```{r x}
x <- mtcars$mpg
```

# Example

Now let's see the function in use.

```{r ex}
cum_skewness(x)
```

Let's plot it out.

```{r plt}
plot(cum_skewness(x), type = "l")
```
