---
title: "Unveiling the Magic of `dcast` Function in R's data.table Package"
author: "Steven P. Sanderson II, MPH"
date: "2024-02-26"
categories: [code, rtip, operations]
---

Here is a draft blog post on using the dcast function from the data.table package in R:

# Introduction

The `dcast` function in the data.table package is a handy tool for reshaping data from long to wide format in R. It allows you to easily aggregate values and pivot dataframes. In this post, we'll explore how to use dcast through several examples.

# What is dcast?

The `dcast` stands for "data cast" and it works similarly to reshape2's `dcast` function. The key difference is that `dcast` from data.table is much faster as it works directly on data.table objects and leverages fast data.table syntax.

To use dcast, you need to have the data.table package installed:

```{r}
library(data.table)
```

The main arguments for dcast are:

- `data`: the data.table or data.frame to reshape
- `formula`: specifies columns to use, in the form `value.var ~ key1 + key2` 
- `fun.aggregate`: the aggregation function to apply (default is length)

## Reshaping from Long to Wide

Let's walk through an example with the mtcars dataset. First we convert mtcars to a data.table:

```{r}
dt <- as.data.table(mtcars)
```

Say we want to reshape the data from long to wide, aggregating the hp values by cyl. We can use dcast:

```{r}
dcast(dt, cyl ~ ., value.var="hp", fun.aggregate=mean)
```

This aggregates the hp by cyl, casting the other columns as identifiers. The result is a table with one row per cyl, and columns for mean hp and all other variables.

## Aggregating Multiple Columns 

You can also aggregate multiple value columns in one call. Let's add aggregating disp by the mean:

```{r}
dcast(dt, cyl ~ ., value.var=c("hp", "disp"), fun.aggregate=mean)
```

Now we have mean hp and mean disp aggregated by cyl in the wide format.

## Using Multiple Formulas

Another common operation is aggregating over several formulas separately. For example, aggregating hp by cyl and gear.

We can pass a list of formulas to dcast:

```{r}
dcast(dt, cyl ~ ., value.var="hp", fun.aggregate=mean) + 
  dcast(dt, gear ~ ., value.var="hp", fun.aggregate=mean)
```

This outputs two sets of aggregations, by cyl and gear, in a single wide table.

## Reshaping from Wide to Long

The `melt` function from data.table can reshape from wide to long format. For example:

```{r}
melt(dt, id.vars = "cyl", measure.vars = c("hp", "disp"))
```

This melts the data to long form based on the id and measure columns.

## Additional Tips

- Use `fun.aggregate=length` to get counts per group 
- Set `fill=NA` to output NA for combinations without data instead of 0 
- Use `variable.name` to set custom column names

# Wrapping Up

The dcast function provides a fast way to reshape data and aggregate values in R. It's perfect for pivoting dataframes and getting data ready for analysis and visualization. The data.table syntax helps make the reshape very fast and efficient. Give it a try on your datasets! Let me know in the comments if you have any other dcast examples to share.