---
title: "Cumulative Harmonic Mean with {TidyDensity}"
author: "Steven P. Sanderson II, MPH"
date: "2022-11-16"
categories: [code, rtip, tidydensity]
---

# Introduction

There can be times in which you may want to see a cumulative statistic, maybe in this particular case it is the harmonic mean. Well with the [`{TidyDensity}`](https://www.spsanderson.com/TidyDensity/) it is possible with a function called [`chmean()`](https://www.spsanderson.com/TidyDensity/reference/chmean.html)

Let's take a look at the function.

# Function

Here is the function call, it is very simple as it is a vectorized function.

```{r fns, eval=FALSE}
chmean(.x)
```

The only argument you provide to this function is a numeric vector. Let's take a quick look at the construction of the function.

```{r fns_c, eval=FALSE}
chmean <- function(.x) {
  1 / (cumsum(1 / .x))
}
```

# Examples

Let's take a look at an example.

```{r example, message=FALSE, warning=FALSE}
library(TidyDensity)
library(dplyr)

x <- mtcars$mpg

chmean(x)

mtcars %>%
  select(mpg) %>%
  mutate(cum_har_mean = chmean(mpg)) %>%
  head(10)
```

Voila!