---
title: "Unlocking the Power of Administrative Data with healthyR.data"
author: "Steven P. Sanderson II, MPH"
date: "2024-07-01"
categories: [code, rtip, heathyrdata]
toc: TRUE
---

# Introduction

Hello, fellow data enthusiasts! Today, I’m excited to share insights into the **{ healthyR.data }** package, an essential tool in the healthyverse that will streamline your data exploration and testing processes. Whether you're a seasoned data scientist or just starting out in data analytics, this package is designed to be valuable for everyone.

## What is { healthyR.data }?

The **{ healthyR.data }** package serves two primary purposes: providing a robust administrative data set for testing functions in the **{ healthyR }** package and facilitating the download of important data from the Centers for Medicare and Medicaid Services (CMS), a division of the Department of Health and Human Services (HHS). This package is your resource for comprehensive data that can enhance your analytical capabilities and simplify your testing procedures.

## Key Features

### 1. Comprehensive Administrative Data Set

One of the main features of **{ healthyR.data }** is its extensive administrative data set. This data set is carefully curated to include a variety of scenarios and variables commonly found in healthcare data analysis. This makes it an excellent tool for testing the functions of the **{ healthyR }** package, ensuring your analytical methods are reliable and effective.

### 2. CMS Data Access

In addition to its built-in data set, **{ healthyR.data }** allows you to download data directly from CMS. This feature is especially useful for healthcare analysts who need up-to-date and detailed data for their analyses. With **{ healthyR.data }**, you can easily access a wealth of information to drive insightful analysis.

## Getting Started

To start using **{ healthyR.data }**, you can install it from CRAN with the following command:

```r
install.packages("healthyR.data")
```

Once installed, load the package with:

```r
library(healthyR.data)
```

## Functions

Load the libraries:

```{r message = FALSE, warning = FALSE}
library(healthyR.data)
library(tidyverse)
library(DT)
```

```{r}
# Functions and their arguments for healthyR

pat <- c("%>%",":=","as_label","as_name","enquo","enquos","expr",
         "sym","syms","healthyR_data")

tibble(fns = ls.str("package:healthyR.data")) |>
  filter(!fns %in% pat) |>
  mutate(params = purrr::map(fns, formalArgs)) |> 
  group_by(fns) |> 
  mutate(func_with_params = toString(params)) |>
  mutate(
    func_with_params = ifelse(
      str_detect(
        func_with_params, "\\("), 
      paste0(fns, func_with_params), 
      paste0(fns, "(", func_with_params, ")")
    )) |>
  select(fns, func_with_params) |>
  mutate(fns = as.factor(fns)) |>
  datatable(
    #class = 'cell-boarder-stripe',
    colnames = c("Function", "Full Call"),
    options = list(
      autowidth = TRUE,
      pageLength = 10
    )
  )
```


## Using the Administrative Data Set

The administrative data set included in **{ healthyR.data }** is ready for your analytical projects. Here’s a quick example:

```r
# Load the healthyR.data package
library(healthyR.data)

# Explore the dataset
data("healthyR_data")
head(healthyR_data)
```

This will give you a look at the data and its structure, providing a strong foundation for your analysis.

## Downloading CMS Data

Accessing CMS data is simple with **{ healthyR.data }**. The package includes functions that allow you to download various datasets directly from the CMS website. Here’s how:

```r
# Download CMS data
cms_data <- get_cms_meta_data()

# View the downloaded data
head(cms_data)
```

This function fetches the latest data from CMS, ensuring your analyses are based on current information.

## Latest Updates and Features

The **{ healthyR.data }** package is continually updated, with new features and improvements added regularly. The latest version, 1.1.0, includes several enhancements that make the package even more powerful and user-friendly. For a detailed overview of the latest updates, check out the [NEWS](https://www.spsanderson.com/healthyR.data/news/index.html#healthyrdata-110) section.

## Conclusion

In summary, **{ healthyR.data }** is a versatile package that provides essential tools for healthcare data analysis. Whether you're testing functions from the **{ healthyR }** package or downloading the latest CMS data, **{ healthyR.data }** has you covered. I encourage you to download the package and explore how it can enhance your analytical projects.

Happy coding, and may your data always be insightful!

---

For more information and detailed documentation, visit the [reference page](https://www.spsanderson.com/healthyR.data/reference/index.html). Stay tuned for more updates and tips on how to get the most out of the healthyverse packages!