---
title: "Multinomial Distribution in R"
author: "Steven P. Sanderson II, MPH"
date: "2023-10-31"
categories: [rtip, distribution]
---

# Introduction

The multinomial distribution is a probability distribution that describes the probability of obtaining a specific number of counts for k different outcomes, when each outcome has a fixed probability of occurring.

In R, we can use the `rmultinom()` function to simulate random samples from a multinomial distribution, and the `dmultinom()` function to calculate the probability of a specific outcome.

# Examples

## Example 1

Suppose we have a fair die, and we want to simulate rolling the die 10 times. We can use the `rmultinom()` function to do this as follows:

```{r}
# Simulate rolling a fair die 10 times
die_rolls <- rmultinom(
  n = 10, size = 1, 
  prob = c(1/6, 1/6, 1/6, 1/6, 1/6, 1/6)
  )

# Print the results
print(die_rolls)
```

## Example 2

Suppose we want to calculate the probability of getting exactly two ones, two threes, two fours, two fives, and two sixes when rolling a fair die 10 times. We can use the `dmultinom()` function to do this as follows:

```{r}
# Calculate the probability of getting exactly two ones, two threes, two fours, two fives, and two sixes when rolling a fair die 10 times
probability <- dmultinom(
  x = c(2, 2, 2, 2, 2), 
  size = 10, 
  prob = c(1/6, 1/6, 1/6, 1/6, 1/6)
  )

# Print the result
print(probability)
```

## Try it on your own!

I encourage readers to try using the `rmultinom()` and `dmultinom()` functions on their own data. For example, you could simulate rolling a die 100 times and see how often each outcome occurs. Or, you could calculate the probability of getting a certain number of heads and tails when flipping a coin 10 times.

Here is an example of how to use the `rmultinom()` function to simulate flipping a coin 10 times and calculate the probability of getting exactly five heads and five tails:

```{r}
# Simulate flipping a coin 10 times
coin_flips <- rmultinom(n = 10, size = 1, prob = c(0.5, 0.5))

# Calculate the probability of getting exactly five heads and five tails
probability <- dmultinom(x = c(5, 5), size = 10, prob = c(0.5, 0.5))

# Print the result
print(probability)
```

I hope this blog post has helped you learn how to use the multinomial distribution in R. Please feel free to leave a comment if you have any questions.