---
title: "Find Skewed Features with {healthyR.ai}"
author: "Steven P. Sanderson II, MPH"
date: "2022-11-14"
categories: [code, rtip, healthyrai, skew]
---

# Introduction

Sometimes we may want to quickly find skewed features in a data set. This is easily achiveable using the `{healthyR.ai}` library. There is a simple function called [`hai_skewed_features()`](https://www.spsanderson.com/healthyR.ai/reference/hai_skewed_features.html). We are going to go over this function today.

# Function

Let's first take a look at the function call.

```{r fns, eval=FALSE}
hai_skewed_features(
  .data, 
  .threshold = 0.6, 
  .drop_keys = NULL
  )
```

Now let's take a look at the arguments that go to the parameters of the function.

*  `.data` - The data.frame/tibble you are passing in.
*  `.threshold` - A level of skewness that indicates where you feel a column should be considered skewed.
*  `.drop_keys` - A c() character vector of columns you do not want passed to the function.


# Example

Here are a couple of examples.

```{r examples, message=FALSE, warning=FALSE}
library(healthyR.ai)

hai_skewed_features(mtcars)

hai_skewed_features(mtcars, .drop_keys = "hp")
```