---
title: "Filtering Rows in R Where Column Value is Between Two Values"
author: "Steven P. Sanderson II, MPH"
date: "2024-03-01"
categories: [code, rtip, operations]
---

# Introduction

Filtering data frames in R is a common task in data analysis. Often we want to subset a data frame to only keep rows that meet certain criteria. A useful filtering technique is keeping rows where a column value falls between two specified values. 

In this post, we'll walk through how to filter rows in R where a column value is between two values using base R syntax.

## Filtering with bracket notation

One way to filter rows is by using bracket notation `[]` and specifying a logical vector. 

Let's create a sample data frame:

```{r}
df <- data.frame(
  id = 1:10,
  value = c(5, 3, 6, 9, 2, 4, 7, 1, 8, 10)
)
``` 

We can filter `df` to only keep rows where `value` is between 5 and 8 with:

```{r}
df[df$value >= 5 & df$value <= 8,]
```

This filters for rows where `value` is greater than or equal to 5 `df$value >= 5` AND less than or equal to 8 `df$value <= 8`. The comma after the logical vector tells R to return the filtered rows.

## Filtering with subset()

Another option is using the `subset()` function:

```{r}
subset(df, value >= 5 & value <= 8)
```

`subset()` takes a data frame as the first argument, then a logical expression similar to the bracket notation.

## Additional examples

We can filter on different columns and value ranges:

```{r}
# id between 3 and 7
df[df$id >= 3 & df$id <= 7,] 

# value less than 5
subset(df, value < 5)
```

It's also possible to filter rows outside a range by flipping the logical operators:

```{r}
# id NOT between 3 and 7
df[!(df$id >= 3 & df$id <= 7),]

# value greater than 5  
subset(df, value > 5) 
```

## Summary

Filtering data frames where a column is between two values is straightforward in R. The key steps are:

- Use bracket notation `df[logical,]` or `subset(df, logical)`
- Create a logical expression with `&` and `>=`, `<=` operators 
- Specify the column name and range of values to filter between

I encourage you to try filtering data frames on your own! Subsetting by logical expressions is an important skill for efficient R programming.