---
title: "Cumulative Variance"
author: "Steven P. Sanderson II, MPH"
date: "2022-10-26"
categories: [code, rtip, healthyrai]
---

# Introduction

Sometimes you may be working with a time series or some process data and you will want to make a control chart. This is simple to do with the `{healthyR.ai}` package.

If you do not already have it, then you can follow the simple code below to get the latest version.

## Installation

You can install the released version of healthyR.ai from CRAN with:

```{r cran_version, eval=FALSE}
install.packages("healthyR.ai")
```

And the development version from GitHub with:

```{r dev_version, eval=FALSE}
# install.packages("devtools")
devtools::install_github("spsanderson/healthyR.ai")
```

Now that we have the latest version installed, lets get some data and then use the function.

# Generate Data

Lets generate some fake data below:

```{r fake_data}
data_tbl <- tibble::tibble(
  day = sample(
    c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday"),
    100, TRUE
  ),
  person = sample(c("Tom", "Jane", "Alex"), 100, TRUE),
  count = rbinom(100, 20, ifelse(day == "Friday", .5, .2)),
  date = Sys.Date() - sample.int(100)
)
```

Lets take a look at the `data_tbl`, to do so we will use the `glimpse()` function from `dplyr`

```{r glimpse_data}
dplyr::glimpse(data_tbl)
```