---
title: "Hyperbolic Transform with healthyR.ai"
author: "Steven P. Sanderson II, MPH"
date: "2022-11-08"
categories: [code, rtip, healthyrai]
---

# Introduction

In data modeling there can be instanes where you will want some sort of hyperbolic transformation of your data. In `{healthyR.ai}` this is easy with the use of the [function `hai_hyperbolic_vec()`](https://www.spsanderson.com/healthyR.ai/reference/hai_hyperbolic_vec.html) along with it's corresponding `augment` and `step` functions.

# Function

The function takes in a numeric vector as it's argument and will transform the data with one of the following:

*  `sin`
*  `cos`
*  `tan`
*  `sincos` This will do: `value = sin(x) * cos(x)`

The full function call is:
```{r fns, eval=FALSE}
hai_hyperbolic_vec(.x, .scale_type = c("sin", "cos", "tan", "sincos"))
```

# Example

```{r example, message=FALSE, warning=FALSE}
library(dplyr)
library(healthyR.ai)
library(tidyr)
library(ggplot2)

len_out <- 25
by_unit <- "month"
start_date <- as.Date("2021-01-01")

data_tbl <- tibble(
  date_col = seq.Date(
    from = start_date, 
    length.out = len_out, 
    by = by_unit
    ),
  b = runif(len_out),
  fv_sin = hai_hyperbolic_vec(b, .scale_type = "sin"),
  fv_cos = hai_hyperbolic_vec(b, .scale_type = "cos"),
  fv_sc  = hai_hyperbolic_vec(b, .scale_type = "sincos")
)

data_tbl
```

# Visual 

```{r viz}
data_tbl %>% 
  pivot_longer(cols = -date_col) %>% 
  ggplot(aes(x = date_col, y = value, color = name)) + 
  geom_line() + 
  facet_wrap(~ name, scales = "free") +
  theme_minimal() +
  labs(color = "")
```