---
title: "Fitting a Distribution to Data in R"
author: "Steven P. Sanderson II, MPH"
date: "2023-11-02"
categories: [rtip, distribution]
---

# Introduction

The gamma distribution is a continuous probability distribution that is often used to model waiting times or other positively skewed data. It is a two-parameter distribution, where the shape parameter controls the skewness of the distribution and the scale parameter controls the spread of the distribution.

# Fitting a gamma distribution to a dataset in R

There are two main ways to fit a gamma distribution to a dataset in R:

1. **Maximum likelihood estimation (MLE)**: This method estimates the parameters of the gamma distribution that are most likely to have produced the observed data.
2. **Method of moments:** This method estimates the parameters of the gamma distribution by equating the sample mean and variance to the theoretical mean and variance of the gamma distribution.

**MLE** is the more common and generally more reliable method of fitting a gamma distribution to a dataset. To fit a gamma distribution to a dataset using MLE, we can use the `fitdist()` function from the `fitdistrplus` package.

```{r message=FALSE, warning=FALSE}
# Install the fitdistrplus package if necessary
#install.packages("fitdistrplus")

# Load the fitdistrplus package
library(fitdistrplus)
library(TidyDensity)

set.seed(123)
data <- tidy_gamma(.n = 500)$y

# Fit a gamma distribution to the data
fit <- fitdist(data, distr = "gamma", method = "mle")
```

The `fit` object contains the estimated parameters of the gamma distribution, as well as other information about the fit. We can access the estimated parameters using the `coef()` function. Now the `tidy_gamma()` function from the TidyDensity package comes with a default setting of a `.scale = 0.3` and `shape = 1`. The rate is `1/.scale`, so by default it is 3.33333

```{r}
# Get the estimated parameters of the gamma distribution
coef(fit)
```

Now let's see how that compares to the built in TidyDensity function:

```{r}
util_gamma_param_estimate(data)$parameter_tbl[1,c("shape","scale","shape_ratio")]
```
In the above, the `shape_ratio` is the `rate`

# Try on your own!

I encourage you to try fitting a gamma distribution to your own data. You can use the `fitdistrplus` package in R to fit a gamma distribution to any dataset. Once you have fitted a gamma distribution to your data, you can use the estimated parameters to generate random samples from the gamma distribution or to calculate the probability of observing a particular value.