---
title: "Introduction to Linear Regression in R: Analyzing the mtcars Dataset with lm()"
author: "Steven P. Sanderson II, MPH"
date: "2023-06-15"
categories: [rtip, linear, regression]
---

# Introduction

The `lm()` function in R is used for fitting linear regression models. It stands for "linear model," and it allows you to analyze the relationship between variables and make predictions based on the data.

Let's dive into the parameters of the `lm()` function:

1. `formula`: This is the most important parameter, as it specifies the relationship between the variables. It follows a pattern: `y ~ x1 + x2 + ...`, where `y` is the response variable, and `x1`, `x2`, etc., are the predictor variables. For example, in the `mtcars` dataset, we can use the formula `mpg ~ wt` to predict the miles per gallon (`mpg`) based on the weight (`wt`) of the cars.

2. `data`: This parameter refers to the dataset you want to use for the analysis. In our case, we'll use the `mtcars` dataset that comes with R.

Now, let's see some examples using the `mtcars` dataset

# Examples

Example 1: Simple Linear Regression
```{r warning=FALSE, message=FALSE}
# Fit a linear regression model to predict mpg based on weight
model <- lm(mpg ~ wt, data = mtcars)

# Print the summary of the model
summary(model)
```

Example 2: Multiple Linear Regression
```{r warning=FALSE, message=FALSE}
# Fit a linear regression model to predict mpg based on weight and horsepower
model <- lm(mpg ~ wt + hp, data = mtcars)

# Print the summary of the model
summary(model)
```

Example 3: Include Interaction Term
```{r warning=FALSE, message=FALSE}
# Fit a linear regression model to predict mpg based on weight, horsepower, and their interaction
model <- lm(mpg ~ wt + hp + wt:hp, data = mtcars)

# Print the summary of the model
summary(model)
```

These examples demonstrate how to use the `lm()` function with different sets of predictor variables. After fitting the model, you can use the `summary()` function to get detailed information about the regression results, including coefficients, p-values, and R-squared values.

I encourage you to try running these examples and explore different variables in the `mtcars` dataset. Feel free to modify the formulas and experiment with additional parameters to deepen your understanding of linear regression modeling in R!