---
title: "The Argument Matcher: A Function for Selecting the Right Arguments {tidyAML}"
author: "Steven P. Sanderson II, MPH"
date: "2023-02-03"
categories: [code, rtip, list, tidyaml, tidymodels]
---

# Introduction

I am working on finishing up a few things with my new R package [`{tidyAML}`](https://www.spsanderson.com/tidyAML/) before I release it to CRAN. One of those things is the ability of a user to build a model using a command that might be something like `generate_model()`. One of the things that is necessary to do is to match the function arguments from the `generate_model()` to the actual __parsnip__ call.

This is where and argument matcher of sorts may come in handy. I am doing this because it will take one most step of abstraction away, and instead of say calling `linear_reg()` or `mars()` or something like that, you can just instead use `generate_model()` and type in your engine or the `parsnip` function call there.

Now I am not one hundred percent certain that I'll actually implement this or not, but the exercise was fun enough that I decided to share it. So let's get into it.

# Function

Here is the current state of the function.

```{r}
argument_matcher <- function(.f = "linear_reg", .args = list()){
  
  # TidyEval ----
  fns <- as.character(.f)
  
  fns_args <- formalArgs(fns)
  fns_args_list <- as.list(fns_args)
  names(fns_args_list) <- fns_args
  
  arg_list <- .args
  arg_list_names <- unique(names(arg_list))
  
  l <- list(arg_list, fns_args_list)
  
  arg_idx <- which(arg_list_names %in% fns_args_list)
  bad_arg_idx <- which(!arg_list_names %in% fns_args_list)
  
  bad_args <- arg_list[bad_arg_idx]
  bad_arg_names <- unique(names(bad_args))
  
  final_args <- arg_list[arg_idx]
  
  # Return ----
  if (length(bad_arg_names > 0)){
    rlang::inform(
      message = paste0("bad arguments passed: ", bad_arg_names),
      use_cli_format = TRUE
    )
  }

  return(final_args)
}
```

When working with R functions, it's not uncommon to encounter a situation where you need to pass arguments to another function. This can be especially challenging when the arguments are not properly matched. Fortunately, the `argument_matcher` function provides an elegant solution to this problem.

The `argument_matcher` function takes two arguments: `.f` and `.args`. The `.f` argument is a string that specifies the name of the function you want to pass arguments to, while the `.args` argument is a list that contains the arguments you want to pass to the specified function.

The `argument_matcher` function first uses the `formalArg`s function to extract the formal arguments of the specified function and store them in `fns_args`. The names of the formal arguments are then used to create a list, `fns_args_list`.

Next, the function extracts the names of the arguments in `.arg`s and stores them in `arg_list_names`. It then checks if the names of the arguments in `.args` match the names of the formal arguments of the specified function, and stores the matching arguments in `final_args`. Any arguments that don't match the formal arguments are stored in bad_args, and a warning message is printed indicating that bad arguments were passed.

The final step is to return the `final_args` list, which contains only the arguments that match the formal arguments of the specified function.

In conclusion, the `argument_matcher` function is a useful tool for ensuring that arguments are properly matched when passed to another function. Whether you're working with linear regression models or any other type of function, the `argument_matche`r function will help you select the right arguments and avoid common errors.

# Example

Let's see a simple example.

```{r warning=FALSE}
suppressPackageStartupMessages(library(tidymodels))

argument_matcher(
  .args = list(
    mode = "regression", 
    engine = "lm",
    cost = 0.5,
    trees = 1, 
    mtry = 1
    )
  )
```

Voila!