---
title: "Unveiling the Smooth Operator: Rolling Averages in R"
author: "Steven P. Sanderson II, MPH"
date: "2024-01-05"
categories: [code, rtip, timeseries]
---

# Introduction

Ever felt those data points were a bit too jittery? Smoothing out trends and revealing underlying patterns is a breeze with rolling averages in R. Ready to roll? Let's dive in!

# Rolling with the 'zoo'

Meet the 'zoo' package, your trusty companion for time series data wrangling. It's got a handy function called 'rollmean' that handles those rolling averages with ease.

# Installing and Loading

```{r message = FALSE, warning = FALSE}
# install.packages("zoo")  # Grab it if you haven't already
library(zoo)  # Bring it into your workspace
```

# Example

## Creating a Simple Time Series

```{r}
set.seed(123)  # Set seed for reproducibility (optional
# Let's imagine some daily sales data
sales <- trunc(runif(112, min = 100, max = 500))  # Generate some random sales
days <- as.Date(1:112, origin = "2022-12-31")  # Add some dates!
data_zoo <- zoo(sales, days)  # Convert to a zoo object
```

## Calculating Rolling Averages

```{r}
# Say we want a 7-day rolling average:
rolling_avg7 <- rollmean(data_zoo, k = 7)
rolling_avg7_left <- rollmean(data_zoo, k = 7, align = "left")
rolling_avg7_right <- rollmean(data_zoo, k = 7, align = "right")

# How about a 28-day one?
rolling_avg28 <- rollmean(data_zoo, k = 28)
rolling_avg28_left <- rollmean(data_zoo, k = 28, align = "left")
rolling_avg28_right <- rollmean(data_zoo, k = 28, align = "right")
```

## Visualizing the Smoothness

```{r}
plot(data_zoo, type = "l", col = "black", lwd = 1, ylab = "Sales")
lines(rolling_avg7, col = "red", lwd = 2, lty = 2)
lines(rolling_avg7_left, col = "green", lwd = 2, lty = 2)
lines(rolling_avg7_right, col = "orange", lwd = 2, lty = 2)
legend(
  "bottomleft", 
  legend = c(
    "Original Data", "7-day Avg", "7-day Avg (left-aligned)", 
    "7-day Avg (right-aligned)"
    ),
  col = c("black", "red", "green", "orange"), 
  lwd = 1, lty = 1:2,
  cex = 0.628
  )
```

```{r}
plot(data_zoo, type = "l", col = "black", lwd = 1, ylab = "Sales")
lines(rolling_avg28, col = "green", lwd = 2, lty = 2)
lines(rolling_avg28_left, col = "steelblue", lwd = 2, lty = 2)
lines(rolling_avg28_right, col = "brown", lwd = 2, lty = 2)
legend(
  "bottomleft", 
  legend = c(
    "Original Data", "28-day Avg", "28-day Avg (left-aligned)", 
    "28-day Avg (right-aligned)"
    ),
  col = c("black", "green", "steelblue", "brown"), 
  lwd = 1, lty = 1:2,
  cex = 0.628
  )
```

# Experimenting and Interpreting

Play with different 'k' values to see how they affect the smoothness. Remember, larger 'k' means more smoothing, but potential loss of detail.

# Your Turn to Roll!

Grab your data and start exploring rolling averages! It's a powerful tool to uncover hidden patterns and trends. Share your discoveries and join the rolling conversation!
