---
title: "Distribution Statistics with {TidyDensity}"
author: "Steven P. Sanderson II, MPH"
date: "2022-12-21"
categories: [code, rtip, tidydensity]
---

# Introduction

If you're working with statistical distributions in R, you may be interested in the [`{TidyDensity}`](https://www.spsanderson.com/TidyDensity/) package. This package provides a set of functions for creating, manipulating, and visualizing probability distributions in a tidy format. One of these functions is [`tidy_chisquare()`](https://www.spsanderson.com/TidyDensity/reference/tidy_chisquare.html), which allows you to create a chi-square distribution with a specified number of degrees of freedom and a non-centrality parameter.

Once you've created a chi-square distribution using `tidy_chisquare()`, you may want to get some summary statistics about the distribution. This is where the [`util_chisquare_stats_tbl()`](https://www.spsanderson.com/TidyDensity/reference/util_chisquare_stats_tbl.html) function comes in handy. This function takes a chi-square distribution (created with `tidy_chisquare()`) as input and returns a tibble with several statistics about the distribution.

Some of the statistics included in the table are:

*  Mean: The mean of the chi-square distribution, also known as the expected value.
*  Variance: The variance of the chi-square distribution, which is a measure of how spread out the data is.
*  Skewness: The skewness of the chi-square distribution, which is a measure of the symmetry of the data.
*  Kurtosis: The kurtosis of the chi-square distribution, which is a measure of the peakedness of the data.

To use the `util_chisquare_stats_tbl()` function, you'll need to install and load the `{TidyDensity}` package first. Then, you can create a chi-square distribution using `tidy_chisquare()` and pass it to `util_chisquare_stats_tbl()` like this:

```{r eval=FALSE}
# install and load TidyDensity
install.packages("TidyDensity")
library(TidyDensity)
library(dplyr)

# create a chi-square distribution with 5 degrees of freedom
distribution <- tidy_chisquare(.df = 5)

# get statistics about the distribution
util_chisquare_stats_tbl(distribution) |>
  glimpse()
```

The output will be a table with the mean, variance, skewness, and kurtosis of the chi-square distribution. These statistics can be useful for understanding the characteristics of the distribution and making statistical inferences.

Overall, the `{TidyDensity}` package is a useful tool for working with statistical distributions in R. The `util_chisquare_stats_tbl()` function is just one of many functions available in the package that can help you analyze and understand your data. Give it a try and see how it can help with your statistical analysis!

# Function

Let's take a look at the full function call.

```{r eval=FALSE, warning=FALSE, message=FALSE}
util_chisquare_stats_tbl(.data)
```

Let's take a look at the arguments that get supplied to the function parameters.

*  `.data` - The data being passed from a tidy_ distribution function.

# Example

Now for a full example with output.

```{r ex1, warning=FALSE, message=FALSE}
library(TidyDensity)
library(dplyr)

tidy_chisquare() %>%
  util_chisquare_stats_tbl() %>%
  glimpse()
```

Voila!