---
title: "How to Get First or Last Day of Month in R with lubridate and base R"
author: "Steven P. Sanderson II, MPH"
date: "2024-02-13"
categories: [code, rtip, timeseries]
---

# Introduction

When working with dates in R, you'll often need to find the first or last day of the current month or any given month. There are a couple easy ways to do this using the lubridate package and base R functions. In this post, I'll show you how.

# Examples

## Using lubridate

The lubridate package makes working with dates in R much easier. It has a number of helper functions for manipulating and extracting info from Date and POSIXct objects.

To get the first day of the current month, you can use `floor_date()` and pass it the current date:

```{r message=FALSE, warning=FALSE}
library(lubridate)

today <- Sys.Date() # or Sys.time() for POSIXct
first_day <- floor_date(today, unit = "month")
first_day
```

This will return a Date object with the first day of the month.

To get the last day, use `ceiling_date()` instead:

```{r message=FALSE, warning=FALSE}
last_day <- ceiling_date(today, unit = "month") - days(1)
last_day
```

You can also pass any Date object to these functions to get the first or last day of that month:

```{r}
date <- ymd("2023-06-15")
floor_date(date, "month") # 2023-06-01
ceiling_date(date, "month") - days(1) # 2023-06-30
```

The lubridate functions make this really easy!

## Base R Methods

You can also get the first and last day of month using just base R functions.

For the first day, use `as.Date()` with `format()` and pass it the year, month, and day 1:

```{r}
first_day <- as.Date(format(today, "%Y-%m-01"))
first_day
```

For the last day, we can use `0` as the day which will give the last day of the month:

```{r}
last_day <- as.Date((format(today + months(1), "%Y-%m-01")))-1
last_day
```

A bit more work than lubridate, but good to know you can do this with just base R.

I hope this helps you easily get the first and last day of the month in your own date analyses in R! Let me know if you have any other questions.