---
title: "How to Extract Substring Starting from the End of a String in R"
author: "Steven P. Sanderson II, MPH"
date: "2024-07-16"
categories: [code, rtip, operations, strings, regex, stringr, stringi]
toc: TRUE
---

# Introduction

Hey useR's! Today, we're going to discuss a neat trick: extracting substrings starting from the end of a string. We'll cover how to achieve this using base R, `stringr`, and `stringi`. By the end of this post, you'll have several tools in your R toolbox for string manipulation. Let's get started!

# Extracting Substring from the End of a String

## Using Base R

First up, let's use base R functions to extract substrings from the end of a string. The `substr` function is your friend here.

Here's a simple example:

```{r}
# Define a string
my_string <- "Hello, world!"

# Extract the last 6 characters
substring_from_end <- substr(my_string, nchar(my_string) - 5, nchar(my_string))

# Print the result
print(substring_from_end)
```

**Explanation:**

-   `nchar(my_string)` returns the total number of characters in `my_string`.
-   `nchar(my_string) - 5` calculates the starting position of the substring, counting from the end.
-   `substr(my_string, start, stop)` extracts the substring from the start position to the stop position.

## Using `stringr`

The `stringr` package makes string manipulation more straightforward and readable. We'll use the `str_sub` function for this task.

First, install and load the `stringr` package if you haven't already:

```{r message=FALSE, warning=FALSE}
#install.packages("stringr")
library(stringr)
```

Now, let's extract a substring from the end:

```{r}
# Define a string
my_string <- "Hello, world!"

# Extract the last 6 characters using stringr
substring_from_end <- str_sub(my_string, -6, -1)

# Print the result
print(substring_from_end)
```

**Explanation:**

-   `str_sub(my_string, start, end)` extracts the substring from the start to the end position.
-   Negative indices in `str_sub` count from the end of the string. So `-6` refers to the sixth character from the end, and `-1` refers to the last character.

## Using `stringi`

The `stringi` package is another powerful tool for string manipulation. We'll use the `stri_sub` function here.

First, install and load the `stringi` package:

```{r message=FALSE, warning=FALSE}
#install.packages("stringi")
library(stringi)
```

Let's extract our substring:

```{r}
# Define a string
my_string <- "Hello, world!"

# Extract the last 6 characters using stringi
substring_from_end <- stri_sub(my_string, from = -6, to = -1)

# Print the result
print(substring_from_end)
```

**Explanation:**

-   `stri_sub(my_string, from, to)` works similarly to `str_sub`, using `from` and `to` parameters to define the start and end positions.
-   Negative values count from the end of the string.

# Try It Yourself!

Now it's your turn! Try these methods on your own strings. Here are a few ideas to get you started:

-   Extract the last 3 characters of your name.
-   Extract the domain from an email address.
-   Experiment with different lengths and positions.

# Conclusion

We’ve explored how to extract substrings from the end of a string using base R, `stringr`, and `stringi`. Each method has its own charm, so choose the one that fits your coding style best. String manipulation is a crucial skill in data cleaning and text analysis, so keep practicing and experimenting.

---

Happy coding! 🚀