---
title: "Unveiling Data Distribution Patterns with `stripchart()` in R"
author: "Steven P. Sanderson II, MPH"
date: "2023-08-22"
categories: [rtip, viz]
---

# Introdution

Data visualization is a powerful tool that allows us to uncover patterns and insights within datasets. One such tool in the R programming arsenal is the `stripchart()` function. If you're looking to reveal distribution patterns in your data with style and simplicity, then this function might just become your new best friend. In this blog post, we'll dive into the world of `stripchart()`, exploring its syntax, uses, and providing you with hands-on examples to master its application.

# Understanding the Syntax

The syntax of the `stripchart()` function is quite straightforward. Here's the basic structure:

```r
stripchart(x, method = "stack", vertical = TRUE, ...)
```

- `x`: This is the input data that you want to visualize. It can be a vector, a numeric matrix, or a data frame.
- `method`: Specifies how the data points are plotted along the y-axis. Options include `"stack"` (default), `"jitter"`, `"overplot"`, and `"overplot"`.
- `vertical`: If set to `TRUE` (default), the chart will be vertical; if set to `FALSE`, it will be horizontal.
- `...`: Additional graphical parameters that can be passed to customize the appearance of the chart.

# Examples

The `stripchart()` function finds its utility in various scenarios, including:

## Example 1. Comparing Distributions

Let's say you have two datasets containing exam scores of students from different schools. You can use `stripchart()` to visually compare their distributions:

```{r}
# Sample data
school_A <- c(70, 72, 75, 78, 80, 85, 88)
school_B <- c(65, 68, 70, 73, 75, 80, 85, 90)

# Create a stripchart
stripchart(list(School_A = school_A, School_B = school_B),
           vertical = FALSE, method = "jitter",
           main = "Exam Score Distributions",
           xlab = "Score", ylab = "School")
```

In this example, we're using the `"jitter"` method to spread out the points along the y-axis, making it easier to see the density of scores.

## Example 2. Visualizing Data Points

Imagine you have a dataset with the heights of individuals. You can use `stripchart()` to visualize each individual's height as a data point:

```{r}
# Sample data
heights <- c(160, 170, 175, 155, 180, 165, 172, 158, 185)

# Create a stripchart
stripchart(heights, method = "overplot",
           main = "Individual Heights",
           xlab = "Height (cm)")
```

In this case, the `"overplot"` method allows us to see individual data points that might overlap.

## Example 3. Categorical Data Comparison

Suppose you have a dataset of employees' years of service. You can use `stripchart()` to compare the years of service among different departments:

```{r}
# Sample data
hr_dept <- c(2, 3, 4, 2, 5, 3)
tech_dept <- c(1, 2, 1, 3, 2, 4, 2)

# Create a stripchart
stripchart(list(HR = hr_dept, Tech = tech_dept),
           vertical = FALSE, method = "stack",
           main = "Years of Service by Department",
           xlab = "Years", ylab = "Department")
```

The `"divide"` method segments data points based on the provided categories.

## Example 4. All three methods in one

Now let's see what all three methods show for the same data set. We will place them all on one plot.

```{r}
x <- rnorm(100)

par(mfrow = c(2, 2))
# Create a stripchart of the heights of 100 randomly generated people
stripchart(x, method = "overplot", main = "Overplot")

# Create a stripchart of the heights of 100 people, jittering the points to prevent overlapping
stripchart(x, method = "jitter", main = "Jitter")

# Create a stripchart of the heights of 100 people, stacking the points on top of each other
stripchart(x, method = "stack", main = "Stack")

par(mfrow = c(1, 1))
```

# Empower Yourself with `stripchart()`

The `stripchart()` function is a fantastic tool for visualizing data distributions, making comparisons, and spotting patterns. It's a simple yet effective way to represent your data graphically. So, don't hesitate to roll up your sleeves and give it a try. The world of data visualization is at your fingertips, waiting for you to unveil its secrets with the power of R's `stripchart()` function. Happy coding!

Remember, the examples we've explored here are just the beginning. Feel free to experiment, tweak parameters, and adapt the function to your specific needs. The more you practice, the more confident you'll become in using this powerful visualization tool. Your data has stories to tell, and `stripchart()` is here to help you tell them vividly and creatively. Enjoy your data exploration journey!