---
title: "R Functions for Getting Objects"
author: "Steven P. Sanderson II, MPH"
date: "2023-08-01"
categories: [rtip]
---

# Introduction

Welcome, fellow programmers, to this exciting journey into the world of R functions! Today, we'll explore four powerful functions: `get()`, `get0()`, `dynGet()`, and `mget()`. These functions may sound mysterious, but fear not; we'll demystify them together and see how they can be incredibly handy tools in your R toolkit. So, let's dive in!

# The get() Function

The `get()` function is a versatile and often overlooked gem in R. Its primary purpose is to retrieve the value of a variable stored in an environment by specifying its name as a character string. The syntax of the `get()` function is straightforward:

``` r
get(x, pos = parent.frame())
```

## Example get()

Let's say you have a variable named `my_variable` stored somewhere in your R environment, and you want to access its value using the `get()` function:

```{r}
# Sample variable in the environment
my_variable <- 42

# Using get() to retrieve the value
result <- get("my_variable")
result
```

## Explanation of get()

In the example above, we used `get("my_variable")` to access the value of the variable `my_variable`. The function returned the value `42`, which was stored in the variable.

# The get0() Function

The `get0()` function is closely related to `get()`, but it has a subtle difference. It allows for the passing of an error message via the `ifnotfound` parameters. The syntax of `get0()` is:

``` r
get0(x, envir = pos.to.env(-1L), mode = "any", inherits = TRUE,
     ifnotfound = NULL)
```

## Example get0()

Let's use the same variable `my_variable` as before and see the difference between `get()` and `get0()`:

```{r}
# Sample variable in the environment
my_variable <- 42

# Using get0() to retrieve the variable itself
result <- get0("my_var", ifnotfound = "Does Not Exist")
result
```

## Explanation of get0()

In this example, `get0("my_var")` returned an error message as the variable was not found.

# The dynGet() Function

The `dynGet()` function is similar to `get()`, but it searches for the variable in a specified environment. The syntax is:

``` r
dynGet(x, ifnotfound = , minframe = 1L, inherits = FALSE)
```

## Example dynGet()

Consider a scenario where you have a variable named `num` inside a custom environment, and you want to access it using `dynGet()`:

```{r}
# Create a new environment
custom_env <- new.env()

# Assign a variable inside the custom environment
custom_env$num <- 99

# Using dynGet() to retrieve the value
result_env <- dynGet("num", custom_env)
result_env

result_num <- dynGet("num", custom_env$num)
result_num
```

## Explanation of dynGet()

In this example, we used `dynGet("num", custom_env$num)` to access the value of the variable `num` from the specified `custom_env` environment. The function successfully retrieved the value `99`.

# The mget() Function

The `mget()` function is a workhorse when you want to retrieve multiple variables at once. It takes a vector of variable names as input and returns a named list with the values. The syntax is:

``` r
mget(..., envir = as.environment(-1))
```

## Example of mget()

Let's say we have two variables, `x` and `y`, and we want to retrieve their values using `mget()`:

```{r}
# Sample variables in the environment
x <- 10
y <- 20

# Using mget() to retrieve the values of multiple variables
result <- mget(c("x", "y"))
result # Output: a named list with values: $x [1] 10, $y [1] 20
```

## Explanation of mget()

In this example, we provided the vector `c("x", "y")` to `mget()`, and it returned a named list with the values of both variables `x` and `y`.

# Conclusion

Congratulations on reaching the end of this blog post! We've covered four essential functions in R: `get()`, `get0()`, `dynGet()`, and `mget()`. These functions enable you to access variables and their values efficiently, whether they reside in the global environment or custom environments. We hope you found this information useful and encourage you to try them out in your R projects. Happy coding, and may your R programming journey be filled with success and joy!
