---
title: "How to Create a Bubble Chart in R using ggplot2"
author: "Steven P. Sanderson II, MPH"
date: "2023-10-23"
categories: [rtip, viz, ggplot2]
---

# Introduction

Bubble charts are a great way to visualize data with three dimensions. The size of the bubbles represents a third variable, which can be used to show the importance of that variable or to identify relationships between the three variables.

To create a bubble chart in R using `ggplot2`, you will need to use the `geom_point()` function. This function will plot points on your chart, and you can use the size aesthetic to control the size of the points.

# Getting Started

Before we begin, ensure you have R and `ggplot2` installed. If you don't have `ggplot2`, you can install it with the command:

```R
install.packages("ggplot2")
```
# Examples 

## Example 1: Basic Bubble Chart

Let's start with a simple example using randomly generated data. We'll create a bubble chart that shows the relationship between two variables and represents a third variable using bubble sizes.

```{r ggplot2_bubble_chart, message=FALSE}
# Load ggplot2 library
library(ggplot2)

# Generate random data
set.seed(123)
data <- data.frame(
  x = rnorm(10),
  y = rnorm(10),
  size = runif(10, min = 5, max = 20)
)

# Create a bubble chart
ggplot(data, aes(x, y, size = size)) +
  geom_point() +
  scale_size_continuous(range = c(3, 10)) +
  labs(
    title = "Basic Bubble Chart", 
    x = "X-Axis", 
    y = "Y-Axis",
    size = "Y") +
  theme_minimal()
```

In this example, we create a bubble chart with random data points, where `x` and `y` are the coordinates, and `size` represents the bubble size. The `geom_point()` function is used to add the points, and we adjust the size range using `scale_size_continuous()`.

## Example 2: Customizing Bubble Chart

Now, let's customize our bubble chart further. We'll use a sample dataset to visualize car data, with car names on the bubbles.

```{r ggplot2_bubble_chart2}
# Sample data
cars <- mtcars
cars$name <- rownames(cars)

# Create a bubble chart
ggplot(cars, aes(x = mpg, y = disp, size = hp, label = name)) +
  geom_point() +
  geom_text(vjust = 1, hjust = 1, size = 3) +
  scale_size_continuous(range = c(3, 20)) +
  labs(
    title = "Customized Bubble Chart", 
    x = "Miles per Gallon", 
    y = "Displacement",
    size = "HP") +
  theme_minimal()
```

In this example, we're using the mtcars dataset to create a bubble chart that displays car names using `geom_text()`. The `vjust` and `hjust` parameters control the text placement.

# Other Ways to Use Bubble Charts

Here are a few examples of bubble charts that you can create using ggplot2:

*    A bubble chart showing the relationship between the population, GDP, and land area of different countries.
*    A bubble chart showing the relationship between the sales, marketing budget, and customer satisfaction of different companies.
*    A bubble chart showing the relationship between the temperature, humidity, and wind speed at different locations on a map.


# You Try!

Creating bubble charts in R is not only informative but also fun! Encourage your readers to experiment with their own datasets and customize these examples. The ggplot2 library offers a wealth of possibilities for creating beautiful and insightful visualizations. So, don't hesitate to dive into R and start charting your data with bubbles!

I hope this guide helps you and your readers in creating engaging bubble charts in R using ggplot2. If you have any questions or need further clarification, feel free to ask. Happy coding, Steve!