---
title: "Decoding the Mystery: How to Interpret Regression Output in R Like a Champ"
author: "Steven P. Sanderson II, MPH"
date: "2023-12-14"
categories: [rtip, regression]
---

# Introduction

Ever run an R regression and stared at the output, feeling like you're deciphering an ancient scroll? Fear not, fellow data enthusiasts! Today, we'll crack the code and turn those statistics into meaningful insights.

**Let's grab our trusty R arsenal and set up the scene:**

-   **Dataset:** `mtcars` (a classic car dataset in R)
-   **Regression:** Linear model with `mpg` as the dependent variable (miles per gallon) and all other variables as independent variables (predictors)

# Step 1: Summon the Stats Gods with "summary()"

First, cast your R spell with `summary(lm(mpg ~ ., data = mtcars))`. This incantation conjures a table of coefficients, p-values, and other stats. Don't panic if it looks like a cryptic riddle! We'll break it down:

```{r}
model <- lm(mpg ~ ., data = mtcars)

summary(model)
```

## Coefficients

These tell you how much, on average, the dependent variable changes for a one-unit increase in the corresponding independent variable (holding other variables constant). For example, a coefficient of 0.05 for `cyl` means for every one more cylinder, mpg is expected to increase by 0.05 miles per gallon, on average.

```{r}
model$coefficients
```

## P-values

These whisper secrets about significance. A p-value less than 0.05 would mean the observed relationship between the variable and mpg is unlikely to be due to chance. The following are the individual p-values for each variable:

```{r}
summary(model)$coefficients[, 4]
```

Now the overall p-value for the model:

```{r}
model_p <- function(.model) {
  
  # Get p-values
  fstat <- summary(.model)$fstatistic
  p <- pf(fstat[1], fstat[2], fstat[3], lower.tail = FALSE)
  print(p)
}

model_p(.model = model)
```

# Step 2: Let's Talk Turkey - Interpreting the Numbers

## Coefficients

Think of them as slopes. A positive coefficient means the dependent variable increases with the independent variable. Negative? The opposite! For example, `disp` has a negative coefficient, so bigger engines (larger displacement) tend to have lower mpg.

## P-values

Imagine a courtroom. A low p-value is like a strong witness, convincing you the relationship between the variables is real. High p-values (like for `am`!) are like unreliable witnesses, leaving us unsure.

# Step 3: Zoom Out - The Bigger Picture

## R-squared

This tells you how well the model explains the variation in mpg. A value close to 1 is fantastic, while closer to 0 means the model needs work. In our case, it's not bad, but there's room for improvement.

```{r}
summary(model)$r.squared
```

## Residuals

These are the differences between the actual mpg values and the model's predictions. Analyzing them can reveal hidden patterns and model issues.

```{r}
data.frame(model$residuals)
```

**Bonus Tip:** Visualize the data! Scatter plots and other graphs can make relationships between variables pop.

**Remember:** Interpreting regression output is an art, not a science. Use your domain knowledge, consider the context, and don't hesitate to explore further!

**So next time you face regression output, channel your inner R wizard and remember:**

-   Coefficients whisper about slopes and changes.
-   P-values tell tales of significance, true or false.
-   R-squared unveils the model's explanatory magic.
-   Residuals hold hidden clues, waiting to be discovered.

With these tools in your belt, you'll be interpreting regression output like a pro in no time! Now go forth and conquer the data, fellow R adventurers!

**Note:** This is just a brief example. For a deeper dive, explore specific diagnostics, model selection techniques, and other advanced topics to truly master the art of regression interpretation.
