---
title: "Wrangling Names in R: Your Guide to the make.names() Function"
author: "Steven P. Sanderson II, MPH"
date: "2024-03-11"
categories: [code, rtip, operations]
---

# Introduction

Ever tried to use a number or special character as a name for a variable or column in R, only to be met with an error? R has specific rules for what constitutes a valid name, and the `make.names` function is your knight in shining armor when it comes to wrangling these names into something R understands.

# What is make.names?

Think of `make.names` as a name janitor. It takes a vector of characters (potential names) and ensures they comply with R's naming conventions. These conventions say a valid name:

* Must start with a letter or a dot (".")
* Can only contain letters, numbers, periods, and underscores
* Cannot be a reserved word in R (like `if`, `else`, or `for`)

# How to Use make.names

Using `make.names` is straightforward. You simply provide it with a character vector containing your desired names, and it returns a new vector with valid names. Here's the basic syntax:

```R
new_names <- make.names(old_names)
```

# Making Names Unique (Optional)

By default, `make.names` doesn't guarantee unique names. If you have duplicates, it might just keep them. To ensure unique names, add the `unique = TRUE` argument:

```R
unique_names <- make.names(old_names, unique = TRUE)
```

This will modify duplicate names slightly to make them distinct.

# Examples in Action!

Let's see `make.names` in action with some examples:

```{r}
# Example 1: Fix numeric names
numbers <- c(10, 20, 30)
valid_names <- make.names(numbers)
print(valid_names)
```

In this case, `make.names` prepends an "X" to each number to make them valid names.

```{r}
# Example 2: Handle special characters
special_chars <- c("data#1", "result$", "graph!")
clean_names <- make.names(special_chars)
print(clean_names)
```

Here, `make.names` removes special characters and replaces them with periods (except for "$" which is removed).

# Give it a Try!

R is a playground for exploration. Here are some challenges to try with `make.names`:

1. Create a vector with names containing spaces and underscores. Use `make.names` to see how it handles them.
2. Try using `make.names` on a data frame's column names. What happens?
3. Explore the `unique = TRUE` argument. Can you think of situations where it might be necessary?

Remember, `make.names` is your friend when dealing with non-standard names in R. By understanding its purpose and using it effectively, you can keep your R code clean and error-free. Happy coding!